Computing solution space properties of combinatorial optimization problems via generic tensor networks

### 2022-11-21 paper/paper.tex
update

\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 1cm 0cm 1cm}, clip]{figures/lattices.pdf}
    \caption{The types of graphs used in the case study in \Cref{sec:entropy}.
    The lattice dimensions are $L\times L$. (a) square lattice graphs. (b) square lattice graphs with a filling factor $p=0.8$.
    (c) King's graphs. (d) King's graphs with a filling factor $p=0.8$.}
    \label{fig:lattices}
\end{figure}

\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig5.pdf}
    \caption{Mean entropy for lattice gases on graphs defined in \Cref{fig:lattices}.
    We sampled $1000$ instances for $p=0.8$ lattices and the error bar is too small to be visible.
    The horizontal black dashed lines are for $\lim_{L\rightarrow \infty} F(L,L)^{1/L^2}$ for the corresponding non-disordered square lattice and King's graphs.
    }
    \label{fig:hardsquare}
\end{figure}

\subsection{The overlap gap property\label{sec:overlap-gap}}
With this tool to enumerate or sample configurations, one can understand the structure of the independent set configuration space, such as the optimization landscape for finding the MISs.
One of the known barriers to finding the MIS is the so-called overlap gap property~\cite{Gamarnik2013, Gamarnik2019}.
If the overlap gap property is present, it means every two large independent sets either have a significant intersection or a very small intersection;
it implies that large independent sets are clustered together.
This clustering property has been used to rigorously prove upper bounds on the performance of local search algorithms~\cite{Gamarnik2013, Gamarnik2019}.
To investigate the overlap gap property, we compute pair-wise Hamming distance distributions of large independent sets as they are good indicators of the presence or absence of overlap gap properties.
We inspect two types of graphs that are particularly interesting, the King's graphs with defects and $3$-regular graphs.
It is known that the MIS problem on a general graph can be mapped to the King's graph with defects~\cite{Garey1977,Ebadi2022}. However, it is not clear whether the MIS problem defined on a randomly generated King's graph with defects can have the overlap gap property.
It is known that finding MISs of a $d$-regular graphs has the overlap gap property~\cite{Rahman2017,Gamarnik2021} when both $d$ and the graph sizes are large, but, it is not known whether, for small $d$, e.g.\ for $3$-regular graphs, this statement remains true.
We randomly generated $9$ instances for each category of King's graph at $0.8$ filling with dimensions $20\times 20$ ($320$ vertices) and $3$-regular graphs with $110$ vertices.
At this problem size, there are too many independent sets to fit into any storage, hence we combine the truncated polynomial and sum-product expression tree to directly sample from the target configuration space.
For each instance $G$, we sample $10^4$ pairs of configurations from the independent sets of sizes $\geq \lceil \gamma \times \alpha(G)\rceil$ and show the pair-wise Hamming distance distribution in \Cref{fig:hamming}.
We observe a clear single peak structure at a fixed distance normalized by the MIS size for the King's graphs, indicating the absence of the overlap gap property in a random King's graph at $0.8$ filling.
Since the MIS problem on an arbitrary graph can be mapped to a King's graph at a certain filling, this result is highly nontrivial. It likely implies 
that the King's graphs with defects mapped from hard MIS instances have a very small measure in the total defected King's graph space. In contrast, 
very different pair-wise Hamming distributions are obtained in \Cref{fig:hamming}(b), where we observed the multiple peak structure when the control parameter $\gamma$ is big enough. It indicates the existence of disconnected clusters in the configuration space of the MIS problem on $3$-regular graphs.
We expect this numerical tool can be used to understand this phenomenon better and to further investigate the graph properties and the geometry of the configuration spaces for a variety of graph instances.
\begin{figure} 
    \includegraphics[width=\textwidth, trim={0.0cm 1cm 0.0cm 0cm}, clip]{figures/fig6.pdf}
    \caption{Pairwise Hamming distances distribution for configurations sampled from independent sets with sizes $\geq \lceil\gamma \times \alpha(G)\rceil$.
    In each plot, the $x$-axis is the Hamming distance normalized by the total number of vertices and the $y$-axis is the probability.
    }
    \label{fig:hamming}
\end{figure}

\subsection{Analyzing quantum and classical algorithms for Maximum Independent Set}
\begin{figure} 
    \centering
    \includegraphics[width=.65\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/exp_configurations.pdf}
    \caption{Visualization of experimental outputs of a quantum algorithm for solving the maximum independent set problem~\cite{Ebadi2022}.
    Each vertex represents an independent set, and each edge represents a pair of independent sets that differ by a swap operation or a vertex addition/removal.}
    \label{fig:exp_configuratoins}
\end{figure}
In a recent work, 
the ability to enumerate configurations and compute independence polynomials was critical in understanding the performance of quantum optimization algorithms for the MIS problem on a Rydberg atom quantum computer~\cite{Ebadi2022}. This work focused on 
exploring King's graphs with $0.8$ filling.   The hardest instances for classical simulated annealing could be accurately predicted from the independence polynomial, which gave information about the density of local minima at different independent set sizes.
On the hardest graph instances for simulated annealing studied in the experiment, a high density of local minima were found at independent set sizes of $\alpha(G)-1,$ which the algorithm became trapped in instead of finding the optimal solution of size $\alpha(G)$.
By enumerating the configurations using techniques described in the present work, we found that 
simulated annealing randomly explores the independent sets of size $\alpha(G)-1$ until an optimum solution is found.
Therefore, the large ratio of local to global minima  prevents simulated annealing from efficiently finding an MIS.

@@

\section{Bounding the MIS enumeration space}\label{sec:bounding}
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/masktree.pdf}
    \caption{Bounded enumeration of maximum independent sets. Here, a circle is a tensor, an arrow specifies the execution direction of a function, $\overline A$ is the Boolean mask for $A$ and $\circ$ is the Hadamard (element-wise) multiplication. (a) is the forward pass with tropical algebra (\Cref{eq:tropical}) for computing $\alpha(G)$.
     (b) is the backward pass for computing Boolean gradient masks.
     (c) is the masked tensor network contraction with tropical algebra combined with sets (\Cref{eq:countingtropicalset}) for enumerating configurations.}
     \label{fig:bounding}
\end{figure}

@@

\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig1.pdf}
    \caption{Benchmark results for computing different solution space properties of independent sets of random three-regular graphs with different tensor element types.
    The time in these plots only includes tensor network contraction, without taking into account the contraction order finding and just-in-time compilation time.
    Legends are properties, algebra, and devices that we used in the computation; one can find the corresponding computed solution space property in Table 1 in the main text.
    (a) time and space complexity versus the number of vertices for the benchmarked graphs.
    (b) The computation time for calculating the MIS size and for counting the number of all independent sets (ISs), the number of MISs,
        the number of independent sets having size $\alpha(G)$ and $\alpha(G)-1$, and finding $100$ largest set sizes.
    (c) The computation time for calculating the independence polynomials with different approaches.
    (d) The computation time for configuration enumeration, including single MIS configuration, the enumeration of all independent set configurations, all MIS configurations, all independent sets,
    and all independent set configurations having size $\alpha(G)$ and $\alpha(G)-1$.
    }
    \label{fig:benchmark}
\end{figure}

>>>>

\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 1cm 0cm 1cm}, clip]{lattices.pdf}
    \caption{The types of graphs used in the case study in \Cref{sec:entropy}.
    The lattice dimensions are $L\times L$. (a) square lattice graphs. (b) square lattice graphs with a filling factor $p=0.8$.
    (c) King's graphs. (d) King's graphs with a filling factor $p=0.8$.}
    \label{fig:lattices}
\end{figure}

\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{fig5.pdf}
    \caption{Mean entropy for lattice gases on graphs defined in \Cref{fig:lattices}.
    We sampled $1000$ instances for $p=0.8$ lattices and the error bar is too small to be visible.
    The horizontal black dashed lines are for $\lim_{L\rightarrow \infty} F(L,L)^{1/L^2}$ for the corresponding non-disordered square lattice and King's graphs.
    }
    \label{fig:hardsquare}
\end{figure}

\subsection{The overlap gap property\label{sec:overlap-gap}}
With this tool to enumerate or sample configurations, one can understand the structure of the independent set configuration space, such as the optimization landscape for finding the MISs.
One of the known barriers to finding the MIS is the so-called overlap gap property~\cite{Gamarnik2013, Gamarnik2019}.
If the overlap gap property is present, it means every two large independent sets either have a significant intersection or a very small intersection;
it implies that large independent sets are clustered together.
This clustering property has been used to rigorously prove upper bounds on the performance of local search algorithms~\cite{Gamarnik2013, Gamarnik2019}.
To investigate the overlap gap property, we compute pair-wise Hamming distance distributions of large independent sets as they are good indicators of the presence or absence of overlap gap properties.
We inspect two types of graphs that are particularly interesting, the King's graphs with defects and $3$-regular graphs.
It is known that the MIS problem on a general graph can be mapped to the King's graph with defects~\cite{Garey1977,Ebadi2022}. However, it is not clear whether the MIS problem defined on a randomly generated King's graph with defects can have the overlap gap property.
It is known that finding MISs of a $d$-regular graphs has the overlap gap property~\cite{Rahman2017,Gamarnik2021} when both $d$ and the graph sizes are large, but, it is not known whether, for small $d$, e.g.\ for $3$-regular graphs, this statement remains true.
We randomly generated $9$ instances for each category of King's graph at $0.8$ filling with dimensions $20\times 20$ ($320$ vertices) and $3$-regular graphs with $110$ vertices.
At this problem size, there are too many independent sets to fit into any storage, hence we combine the truncated polynomial and sum-product expression tree to directly sample from the target configuration space.
For each instance $G$, we sample $10^4$ pairs of configurations from the independent sets of sizes $\geq \lceil \gamma \times \alpha(G)\rceil$ and show the pair-wise Hamming distance distribution in \Cref{fig:hamming}.
We observe a clear single peak structure at a fixed distance normalized by the MIS size for the King's graphs, indicating the absence of the overlap gap property in a random King's graph at $0.8$ filling.
Since the MIS problem on an arbitrary graph can be mapped to a King's graph at a certain filling, this result is highly nontrivial. It likely implies 
that the King's graphs with defects mapped from hard MIS instances have a very small measure in the total defected King's graph space. In contrast, 
very different pair-wise Hamming distributions are obtained in \Cref{fig:hamming}(b), where we observed the multiple peak structure when the control parameter $\gamma$ is big enough. It indicates the existence of disconnected clusters in the configuration space of the MIS problem on $3$-regular graphs.
We expect this numerical tool can be used to understand this phenomenon better and to further investigate the graph properties and the geometry of the configuration spaces for a variety of graph instances.
\begin{figure} 
    \includegraphics[width=\textwidth, trim={0.0cm 1cm 0.0cm 0cm}, clip]{fig6.pdf}
    \caption{Pairwise Hamming distances distribution for configurations sampled from independent sets with sizes $\geq \lceil\gamma \times \alpha(G)\rceil$.
    In each plot, the $x$-axis is the Hamming distance normalized by the total number of vertices and the $y$-axis is the probability.
    }
    \label{fig:hamming}
\end{figure}

\subsection{Analyzing quantum and classical algorithms for Maximum Independent Set}
\begin{figure} 
    \centering
    \includegraphics[width=.65\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{exp_configurations.pdf}
    \caption{Visualization of experimental outputs of a quantum algorithm for solving the maximum independent set problem~\cite{Ebadi2022}.
    Each vertex represents an independent set, and each edge represents a pair of independent sets that differ by a swap operation or a vertex addition/removal.}
    \label{fig:exp_configuratoins}
\end{figure}
In a recent work, 
the ability to enumerate configurations and compute independence polynomials was critical in understanding the performance of quantum optimization algorithms for the MIS problem on a Rydberg atom quantum computer~\cite{Ebadi2022}. This work focused on 
exploring King's graphs with $0.8$ filling.   The hardest instances for classical simulated annealing could be accurately predicted from the independence polynomial, which gave information about the density of local minima at different independent set sizes.
On the hardest graph instances for simulated annealing studied in the experiment, a high density of local minima were found at independent set sizes of $\alpha(G)-1,$ which the algorithm became trapped in instead of finding the optimal solution of size $\alpha(G)$.
By enumerating the configurations using techniques described in the present work, we found that 
simulated annealing randomly explores the independent sets of size $\alpha(G)-1$ until an optimum solution is found.
Therefore, the large ratio of local to global minima  prevents simulated annealing from efficiently finding an MIS.

@@

\section{Bounding the MIS enumeration space}\label{sec:bounding}
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{masktree.pdf}
    \caption{Bounded enumeration of maximum independent sets. Here, a circle is a tensor, an arrow specifies the execution direction of a function, $\overline A$ is the Boolean mask for $A$ and $\circ$ is the Hadamard (element-wise) multiplication. (a) is the forward pass with tropical algebra (\Cref{eq:tropical}) for computing $\alpha(G)$.
     (b) is the backward pass for computing Boolean gradient masks.
     (c) is the masked tensor network contraction with tropical algebra combined with sets (\Cref{eq:countingtropicalset}) for enumerating configurations.}
     \label{fig:bounding}
\end{figure}

@@

\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{fig1.pdf}
    \caption{Benchmark results for computing different solution space properties of independent sets of random three-regular graphs with different tensor element types.
    The time in these plots only includes tensor network contraction, without taking into account the contraction order finding and just-in-time compilation time.
    Legends are properties, algebra, and devices that we used in the computation; one can find the corresponding computed solution space property in Table 1 in the main text.
    (a) time and space complexity versus the number of vertices for the benchmarked graphs.
    (b) The computation time for calculating the MIS size and for counting the number of all independent sets (ISs), the number of MISs,
        the number of independent sets having size $\alpha(G)$ and $\alpha(G)-1$, and finding $100$ largest set sizes.
    (c) The computation time for calculating the independence polynomials with different approaches.
    (d) The computation time for configuration enumeration, including single MIS configuration, the enumeration of all independent set configurations, all MIS configurations, all independent sets,
    and all independent set configurations having size $\alpha(G)$ and $\alpha(G)-1$.
    }
    \label{fig:benchmark}
\end{figure}
